Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-13 paper/ms.tex
update ms

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical pathways that produce these objects.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10,Myr to $\approx$1,Gyr timescales. For most known
exoplanets, these processes have already run their course.

Young ($<$1\,Gyr) exoplanets are key to building the timeline for
exoplanet evolution.  Informative individual young systems include
V1298~Tau, an adolescent resonant chain of four transiting puffy
planets that is a likely precusor to the compact multiplanet systems
\citep{David_2019}, and HIP~67522, a Jupiter-sized planet with a
Neptunian mass (\citealt{Rizzuto_2020}; Thao et al.~submitted).
Population-level analyses have similarly suggested differences in the
size distribution of young exoplanets relative to their older
counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q}.  In
this article we will consider planets previously known from transits,
and infer new stellar ages using rotation and lithium.

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr \citep[e.g.~Fig~17
of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
young stars: under this assumption, $\approx$1\% of nearby stars have
$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
currently forming protoplanets \citep{2018A&A...617A..44K}, or of
exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the galactic star
formation rate.

@@

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
most constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
Different sub-groups of the complex may vary in age by $\approx$50\%
(\citealt{Bouma_2022b}; Kerr~et.al.~in preparation).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
subgroup.

@@

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates
of main sequence stars.  Binarity is one example: even
intermediate-separation ($\sim$10\,AU) binaries are biased toward
rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
Metallicity may also be a second-order effect
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
were interested in not only the overall age distribution of the Kepler
field, but also in the reliability of individual ages for young planet
hosting stars.  The quality flagging scheme in Table~\ref{tab:planets}
provides one path toward this end.  The other method is to consider
the region in which we do have overlapping sensitivity between lithium
and rotation -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
for such stars include \ltonegyrhighqconfirmedtwosided\ planets
orbiting ``high-quality'' stars for which the two age indicators are
both constraining and in agreement (with $Q_{\rm star}$ bits zero
through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a precise
lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include only the youngest
  confirmed Kepler planets, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A shell script to decode the $Q_{\rm star}$ quality flag is {\bf \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available online}}.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
%  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{``Confirmed'' planets appear in the machine-readable version before``candidate'' planets.
    The bit quality flags for the rotation-based ages,
    $Q_{\rm star}$, are described in Section~\ref{subsec:flags}.
    Concisely summarized, they are:
    Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
    Bit 1: $\log g$$<$4.2?
    Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
    Bit 3: In KEBC?
    Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
    Bit 5: Confused Kep-Gaia crossmatch?
    Bit 6: Gaia DR3 non-single star?
    Bit 7: RUWE$>$1.4?
    Bit 8: Crowded?
    Bit 9: Far from main-sequence?
    Bit 10: \citetalias{Santos_2021} CP/CB?
    Bit 11: $P_{\rm rot}$ not in the homogeneous
    \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
    As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
    The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
    following meaning.
    Bit 0: Candidate not reliable?
    Bit 1: MES$<$10?
    Bit 2: Grazing?
    For a star to have a high likelihood of being ``reliable for
    gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
    not be raised, and for a planet to be ``reliabe'', we suggested
    $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}

>>>>

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical pathways that produce these objects.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales. For most known
exoplanets, these processes have already run their course.

Young ($<$1\,Gyr) exoplanets represent one means of building the timeline for
exoplanet evolution.  Informative individual young systems include
V1298~Tau, an adolescent resonant chain of four transiting puffy
planets that is a likely precusor to the compact multiplanet systems
\citep{David_2019}, and HIP~67522, a Jupiter-sized planet with a
Neptunian mass (\citealt{Rizzuto_2020}; Thao et al.~submitted).
Population-level analyses have similarly suggested differences in the
size distribution of young exoplanets relative to their older
counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.  In
this article we will consider planets previously known from transits,
and infer new stellar ages using rotation and lithium.

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed from 0--10\,Gyr \citep[e.g.~Fig~17
of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
young stars: under this assumption, $\approx$1\% of nearby stars have
$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
currently forming protoplanets \citep{2018A&A...617A..44K}, and of
exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the galactic star
formation rate.

@@

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
most constraining age measurement from relative isochrone dating of
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
Different sub-groups of the complex may vary in age by $\approx$50\%
(\citealt{Bouma_2022b}; Kerr~et al.,~in preparation).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
subgroup.

@@

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates
of main sequence stars.  Binarity is one example: even
intermediate-separation ($\sim$10\,AU) binaries are biased toward
rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
Metallicity may also be relevant
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
were interested in not only the overall age distribution of the Kepler
field, but also in the reliability of individual ages for young planet
hosting stars.  The quality flagging scheme in Table~\ref{tab:planets}
provides one path toward this end.  The other method is to consider
the cases for which lithium
and rotation overlap in sensitivity -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
for such stars include \ltonegyrhighqconfirmedtwosided\ planets
orbiting ``high-quality'' stars for which the two age indicators are
both constraining and in agreement (with $Q_{\rm star}$ bits zero
through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a precise
lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include only the youngest
  confirmed Kepler planets, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode any given $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
%  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main-sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  Bit 11: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
  The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
  not be raised, and for a planet to be ``reliabe'', we suggested
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}
